---
title: "rbmi: Frequentist and information-anchored inference for reference-based conditional mean imputation"
author: Craig Gower-Page, Alessandro Noci, and Marcel Wolbers
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    number_sections: true
    citation_package: natbib
    base_format: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: true   
linkcolor: blue
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{rbmi: CondMean_InfoAnchored}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

This vignette shows how to obtain information-anchored inference using conditional mean imputation combined with a resampling method such as jackknife or bootstrap as resampling. 

The method follows a proposal by Lu (2021). They suggested to obtain the variance for reference-based imputation methods by decoupling variance estimation from treatment effect estimation. To achieve this, they replace the reference-based imputation by a MAR imputation combined with a delta-adjustment where delta is selected in a data-driven way to match the reference-based estimator. In the following, we show that this can be easily implemented using `rbmi`. 

For more information on the statistical background of this method, please refer to the vignette describing the statistical specifications of the package (`vignette(topic = "stat_specs", package = "rbmi")`). 

## Data and model specification

We use a publicly available example dataset from an antidepressant clinical trial of an active drug versus placebo. The relevant endpoint is the Hamilton 17-item depression rating scale (HAMD17) which was assessed at baseline and at weeks 1, 2, 4, and 6. Study drug discontinuation occurred in 24% of subjects from the active drug and 26% of subjects from placebo. All data after study drug discontinuation are missing and there is a single additional intermittent missing observation.

We consider an imputation model with the mean change from baseline in the HAMD17 score as the outcome (variable CHANGE in the dataset). The following covariates are included in the imputation model: the treatment group (THERAPY), the (categorical) visit (VISIT), treatment-by-visit interactions, the baseline HAMD17 score (BASVAL), and baseline HAMD17 score-by-visit interactions. A common unstructured covariance matrix structure is assumed for both groups. The analysis model is an ANCOVA model with the treatment group as the primary factor and adjustment for the baseline HAMD17 score. For this example, we assume that the imputation strategy after the ICE "study-drug discontinuation" is Jump To Reference (JR) for all subjects and the imputation is based on conditional mean imputation with jackknife as resampling. 

## Draws

We will make use of `rbmi::expand_locf()` to expand the dataset in order to have one row per subject per visit with missing outcomes denoted as `NA`. We will then construct the `data_ice`, `vars` and `method` input arguments to the first core `rbmi` function, `draws()`. For further the details on the specification of these arguments, please refer to the quickstart vignette (`vignette(topic = "stat_specs", package = "rbmi"`) and to the helper functions.

```{r}
library(rbmi)
library(dplyr)

dat <- antidepressant_data

# Use expand_locf to add rows corresponding to visits with missing outcomes to the dataset
dat <- expand_locf(
    dat,
    PATIENT = levels(dat$PATIENT), # expand by PATIENT and VISIT 
    VISIT = levels(dat$VISIT),
    vars = c("BASVAL", "THERAPY"), # fill with LOCF BASVAL and THERAPY
    group = c("PATIENT"),
    order = c("PATIENT", "VISIT")
)

# create data_ice and set the imputation strategy to JR for
# each patient with at least one missing observation
dat_ice <- dat %>% 
    arrange(PATIENT, VISIT) %>% 
    filter(is.na(CHANGE)) %>% 
    group_by(PATIENT) %>% 
    slice(1) %>%
    ungroup() %>% 
    select(PATIENT, VISIT) %>% 
    mutate(strategy = "JR")

# In this dataset, subject 3618 has an intermittent missing values which does not correspond
# to a study drug discontinuation. We therefore remove this subject from `dat_ice`. 
# (In the later imputation step, it will automatically be imputed under the default MAR assumption.)
dat_ice <- dat_ice[-which(dat_ice$PATIENT == 3618),]

# Define the names of key variables in our dataset and
# the covariates included in the imputation model using `set_vars()`
vars <- set_vars(
    outcome = "CHANGE",
    visit = "VISIT",
    subjid = "PATIENT",
    group = "THERAPY",
    covariates = c("BASVAL*VISIT", "THERAPY*VISIT")
)

# Define which imputation method to use (here: conditional mean imputation with jackknife as resampling) 
method <- method_condmean(type = "jackknife")

# Create samples for the imputation parameters by running the draws() function
drawObj <- draws(
    data = dat,
    data_ice = dat_ice,
    vars = vars,
    method = method,
    quiet = TRUE
)
drawObj
```

## Imputation phase including calculation of delta-adjustment

We aim to obtain the information-anchored variance for reference-based imputation methods by decoupling variance estimation from treatment effect estimation. To achieve this, we replace the reference-based imputation by a MAR imputation combined with a delta-adjustment where delta is selected in a data-driven way to match the reference-based estimator. Using rbmi, we first need to perform the imputation under the defined strategy (here JR) and under MAR, separately. Then we access the imputed datasets and we compute the difference between the imputation under JR and the imputation under MAR. This will give us the delta-adjustment that we can add on top of the MAR imputation to match the JR imputation.

```{r}

references <- c("DRUG" = "PLACEBO", "PLACEBO" = "PLACEBO")

# Impute under JR
res_imp <- impute(
      draws = drawObj,
      update_strategy = dat_ice,
      references = references
    )
    
    # Access imputed dataset (index=1 for method_condmean(type = "jackknife"))
    cmi <- extract_imputed_dfs(res_imp, index = 1, idmap = TRUE)[[1]] %>%
      dplyr::select(PATIENT, VISIT, CHANGE) %>%
      rename(y_imp = CHANGE)
    
    # Map back original patients id since rbmi re-code ids to ensure id uniqueness
    cmi <- cmi %>% mutate(PATIENT = factor(recode(PATIENT, !!! attributes(cmi)$idmap), levels = attributes(cmi)$idmap)) 
    
    # Derive conditional mean imputations under MAR
    dat_ice_MAR <- dat_ice %>%
      mutate(strategy = ifelse(strategy %in% c("CIR","CR","JR"), "MAR", strategy))
    
    # Impute under MAR
    impute_obj_MAR <- impute(
      drawObj,
      update_strategy = dat_ice_MAR
    ) # if there are observed post-ICE data, this would give a warning which can be ignored here 
    
    # Take imputed outcome
    cmi_MAR <- extract_imputed_dfs(impute_obj_MAR, index = 1, idmap = TRUE)[[1]] %>%
      dplyr::select(PATIENT, VISIT, CHANGE) %>%
      rename(y_MAR = CHANGE)
    
    # Map back original patients id since rbmi re-code ids to ensure id uniqueness
    cmi_MAR <- cmi_MAR %>% mutate(PATIENT = factor(recode(PATIENT, !!! attributes(cmi_MAR)$idmap), levels = attributes(cmi_MAR)$idmap)) 
    
    # Derive delta adjustment "aligned with ref-based imputation", i.e difference between JR and MAR imputation
    delta_adjust <- full_join(cmi, cmi_MAR, by = c("PATIENT", "VISIT")) %>%
      mutate(delta = y_imp - y_MAR)
```

## Analyse

We can now perform the analysis of the imputed datasets after adding the delta-adjustment. `analyse()` will take as the input argument `imputations` the imputation object corresponding to the MAR imputation (and not the JR imputation!) and will add the delta adjustment `delta` prior to the analysis.

```{r}

    # Set analysis variables using rbmi function "set_vars"
    vars_an <- set_vars(
      group = vars$group,
      visit = vars$visit,
      outcome = vars$outcome,
      covariates = "BASVAL"
    )
    
    # Analyse MAR imputation with derived delta adjustment
    anaObj_MAR_delta <- analyse(
      impute_obj_MAR,
      rbmi::ancova,
      delta = delta_adjust,
      vars = vars_an
    )
```

## Pool

We can finally use the `pool()` function to extract the treatment effect estimate (as well as the estimated marginal means) at each visit and apply the jackknife variance estimator using the analysis of all the imputed leave-one-out samples.

```{r}
  
   poolObj_MAR_delta <- pool(anaObj_MAR_delta)

poolObj_MAR_delta
```
